---
output: github_document
---
   
<!-- README.md is generated from README.Rmd. Please edit that file --> 

```{r, include = FALSE}
knitr::opts_chunk$set(      
  collapse = TRUE,
  warning = FALSE,  
  comment = "#>",
  message = F
)        

```

## {ggprop.test} is a micro *package* to teach the logic of the prop test.

Telling a story with data is a popular idea.  

> The book is meant as a guide to making visualizations that accurately reflect the data, *tell a story*, and look professional. - [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/), Clause Wilke...

> Companies exist that put this idea front and center to their missions: [1. Story Telling with Data](https://www.storytellingwithdata.com/), [2. Building Stories with Data](https://www.cararthompson.com/).

But the focus for 'data storytelling' tends to be on reaching broad audiences, communicating statistical summaries, and compelling complete plots. (See also ['the *glamour* of graphics'](https://www.youtube.com/watch?v=h5cTacaWE6I)).

{ggprop.test}, in contrast, is an attempt to capture the *statistical stories* that are told in *classrooms* all the time.

{ggprop.test} exists to allow instructors and students to engage with the logic of statistical tests and techniques, often presented step-by-step on a chalkboard in a class room, but programmatically! 

Which we don't aim to replace the chalkboard/paper experience but compliment it - indeed it's the inspiration for this project.  See [New approaches to light-weight ‘geom’ (layer) extension]() https://evamaerey.github.io/mytidytuesday/2024-10-29-asa-cowy-fall-2024/asa-cowy-fall-2024.html#22)

This might mean that instead of using the logic a handful of times for a handful of examples, student and instructors might walk through this logic *many* times - becoming not just familiar with the logic, but fluent.

Under the hood, ggplot2 extension is used so that individual concepts can be delivered in both a semantic and visual way. 

{ggprop.test} is an MVP (a minimial viable product/package).  This type of package identifies what is required to deliver functionality, but stops short of putting in all the work that might be required to get a package to CRAN and doesn't adheare to all of [packaging best practices](https://r-pkgs.org/), noting that student feedback and might lead to pretty dramatic changes and rewrites.

Let's have a look at the MVP structure, which does have a lot in common with full-blown CRAN-ready packages:  https://github.com/EvaMaeRey/ggprop.test

```{r}
fs::dir_tree()
```

---

# Student start!

> Motivating question:  For survey of 161 individuals on willingness to serve as organ doners in the case of an accident, is there evidence that responses rate differs from a 50/50 split, when 53 individuals respond 'no' and 108 individuals respond 'yes'?

> Does the sample provide statistical evidence that there isn't a 50-50 split in preference for donation?  Or indiffernece (like people are just answering randomly because they are not paying attention or )


{ggprop.test} can be installed as follows...

```{r, eval = F}
library(remotes)
install_github("EvaMaeRey/ggprop.test")
```

```{r}
library(tidyverse)

library(ggprop.test)


donor_data |> 
  pull(decision) |> 
  table()
```




```{r}
library(tidyverse)

isi_donor_url <- "https://www.isi-stats.com/isi/data/prelim/OrganDonor.txt"

donor_data <- read_delim(isi_donor_url) %>%
  select(Choice) %>% 
  dplyr::mutate(decision = ifelse(Choice == "donor", "donor (1)", "not (0)")) %>% 
  dplyr::mutate(decision = fct_rev(decision)) |> 
  sample_frac()

head(donor_data)

dolphin_data <- tribble(~observed,
                         "Correct (1)",
                         "Correct (1)",
                         "Correct (1)",
                         "Correct (1)",
                         "Correct (1)",
                         "Correct (1)",
                         "Not Correct (0)",
                         "Correct (1)",
                         "Correct (1)",
                         "Correct (1)",
                         "Correct (1)",
                         "Correct (1)",
                         "Correct (1)",
                         "Correct (1)",
                         "Correct (1)",
                         "Correct (1)") |> 
  dplyr::mutate(observed = fct_rev(observed))


```



```{r}
usethis::use_data(dolphin_data, overwrite = T)
usethis::use_data(donor_data, overwrite = T)
```




```{r, eval = T, echo = T}
library(ggprop.test)
snapshot <- ggplyr::intercept


# Does the sample provide statistical evidence that there isn't a 50-50 split in preference for donation?
# or indiffernece (like people are just answering randomly because they are not paying attention or )
donor_data |> 
  pull(decision) |> 
  table()


dolphin_data |> 
  pull(observed) |> 
  table()
```




```{r}
donor_data |>
  ggplot() + 
  aes(x = decision) +
  geom_stack() + 
  geom_stack_label() + snapshot("p1") +
  geom_support() + snapshot("p2") +
  geom_prop() + 
  geom_prop_label() + snapshot("p3") +
  stamp_prop() + 
  stamp_prop_label() + snapshot("p4") +
  stamp_eq_norm_prop() +
  geom_normal_prop_null() + snapshot("p5") +
  geom_normal_prop_null_sds() + snapshot("p6")


# Doris and buzz cooperate to get fish
dolphin_data |>
  ggplot() + 
  aes(x = observed) +
  geom_stack() + 
  geom_stack_label() + 
  geom_support() + 
  geom_prop() + 
  geom_prop_label() + 
  stamp_prop() + 
  stamp_prop_label() + 
  stamp_eq_norm_prop() +
  geom_normal_prop_null() + 
  geom_normal_prop_null_sds() + 
  labs(title = "Are Dolphins Buzz and Doris Cooperating?") + 
  labs(caption = "Wait! Are the 'validity conditions' met here to use the normal approximation?")
```

If we discuss each of the snapshot points, we could write something like this:

```{r, fig.width=12, fig.height=12}
library(patchwork)
(p1 + p2) / 
(p3 + p4) /
(p5 + p6)  + 
  patchwork::plot_annotation(
    tag_levels = 1,
    title = "A prop test 'graphical poem' addressing the question: 
    
        For survey of 161 individuals on willingness to serve as 
        organ doners in the case of an accident, is there evidence
        that responses rate differs from a 50/50 split, 
        when 53 individuals respond 'no' and 108 individuals respond 'yes'?
      ",
    subtitle = "1. Draw bar chart (stacks) for each category.
2. Draw theoretical limits for proportion (0 to 1).
3. Calculate proportion and place (also the balancing point for stacks!).
4. Place null (look at the initial question, 'differs from 50/50' in this case)
5. Draw distribution for NULL, where SD = sqrt((p * (1-p))/n)
6. Calc z-score - where does observed prop fit into NULL distributions 
   (i.e. how many standard deviations fit between .5 and .67)"
                             ) &
  ggchalkboard:::theme_chalkboard(base_size = 14) 


```

---

# What's inside and how did we get here?
   
## step 00. cloning statexpress functions

Some convenience functions from {statexpress} are used, because we want this to be a bit more self-contained at this point, so we just clone them for now.  statexpress is evolving and is not on CRAN.

<details>

```{r statexpress}
qlayer <- function (mapping = NULL, data = NULL, geom = ggplot2::GeomPoint, stat = StatIdentity, 
    position = position_identity(), ..., na.rm = FALSE, show.legend = NA, 
    inherit.aes = TRUE) 
{
    ggplot2::layer(data = data, mapping = mapping, geom = geom, 
        stat = stat, position = position, show.legend = show.legend, 
        inherit.aes = inherit.aes, params = rlang::list2(na.rm = na.rm, 
            ...))
}

qstat <- function (compute_group = ggplot2::Stat$compute_group, ...) 
{
    ggplot2::ggproto(NULL, Stat, compute_group = compute_group, 
        ...)
}

qstat_panel <- function (compute_panel, ...) 
{
    ggplot2::ggproto(NULL, Stat, compute_panel = compute_panel, 
        ...)
}


proto_update <- function (`_class`, `_inherit`, default_aes_update = NULL, ...) 
{
    if (!is.null(default_aes_update)) {
        default_aes <- aes(!!!modifyList(`_inherit`$default_aes, 
            default_aes_update))
    }
    ggplot2::ggproto(`_class` = `_class`, `_inherit` = `_inherit`, 
        default_aes = default_aes, ...)
}

qproto_update <- function (`_inherit`, default_aes_update = NULL, ...) 
{
    proto_update(NULL, `_inherit`, default_aes_update = default_aes_update, 
        ...)
}

```




Now let's see the compute...

```{r compute_for_prop_story, echo = T, message=F, warning=F}
library(tidyverse)

# 1. layer stack of bricks
compute_group_bricks <- function(data, scales, width = .2){
  
  data %>% 
    dplyr::mutate(row = row_number()) %>% 
    dplyr::mutate(y = row - .5) %>% 
    dplyr::mutate(width = width)
  
}

# 2. layer label stack with count
compute_group_count <- function(data, scales){
  
  data %>% 
    dplyr::count(x) %>% 
    dplyr::mutate(y = n,
           label = n)
  
}


# 3. layer add x span
compute_balance <- function(data, scales){
  
  data %>% 
    dplyr::summarise(min_x = min(x),
              xend = max(x),
              y = 0,
              yend = 0) %>% 
    dplyr::rename(x = min_x)
  
}


# 4. layer add balancing point 
compute_xmean_at_y0 <- function(data, scales){
  
  data %>% 
    dplyr::summarise(x = mean(x),
              y = 0, 
              label = "^") 
  
}

# 5. layer add balancing point value label
compute_xmean_at_y0_label <- function(data, scales){
  
  data %>% 
    dplyr::summarise(x = mean(x),
              y = 0, 
              label = after_stat(round(x, 2))) 
  
}



# 6. Add 'point' for asserted balancing point (null)
compute_panel_prop_asserted <- function(data, scales, value = .5){
  
  # stamp type layer - so ignore input data
  data.frame(y = 0, 
             x = null,
             label = "^"
             )
  
}

# 6. Add label for asserted balancing point (null)
compute_panel_prop_asserted_label <- function(data, scales, value = .5){
  
  # stamp type layer - so ignor input data
  data.frame(y = 0, 
             x = null,
             label = round(null, 2)
             )
  
}




# 7. normal distribution based on null and n
compute_dnorm_prop <- function(data, scales, null = .5, dist_sds = seq(-3.5, 3.5, by = .1)
){
  
  n <- data |> nrow()
  n_max <- data |> dplyr::count(.by = x) |> dplyr::pull() |> max()

  
  sd = sqrt(null * (1 - null)/n) # sd of the null distribution
  
  q <- dist_sds * sd + null
  
  data.frame(x = q) %>%
    dplyr::mutate(height = dnorm(q, sd = sd, mean = null)) %>%
    dplyr::mutate(height_max = dnorm(0, sd = sd, mean = 0)) %>%
    dplyr::mutate(y = .55*n_max*height/height_max) %>%  # This is a bit fragile...
    dplyr::mutate(xend = x,
           yend = 0) %>% 
    # @teunbrand ggplot2::GeomArea$setup_data() requires a group column. Your panel computation does not preserve groups, but it should.
    dplyr::mutate(group = 1) 
  
}  


# 8. normal distribution mean and sds based on null and n
compute_dnorm_prop_sds <- function(data, scales, null = .5,
  dist_sds = -4:4){
  
  n <- data |> nrow()
  
  n_max <- data |> dplyr::count(.by = x) |> dplyr::pull() |> max()
  
  sd = sqrt(null * (1 - null)/n) # sd of the null distribution
  
  q <- dist_sds * sd + null
  
  data.frame(x = q) %>%
    dplyr::mutate(height = dnorm(q, sd = sd, mean = null)) %>%
    dplyr::mutate(height_max = dnorm(0, sd = sd, mean = 0)) %>%
    dplyr::mutate(y = .55*n_max*height/height_max) %>% # This is a bit fragile...
    dplyr::mutate(xend = x,
           yend = 0)

}  
```

```{r stamp_eq_norm_prop}
GeomTextBig <- ggproto("GeomTextBig", GeomText,
                       default_aes = modifyList(GeomText$default_aes,
                                                aes(size = from_theme(fontsize))
                                                ))


#' @export
stamp_eq_norm_prop <- function(x = I(.125),
    y = I(.8), ...){
  
  annotate(
    "text",
    x = x,
    y = y,
    label = latex2exp::TeX("sd = \\sqrt{\\frac{p*(1-p)}{n}}", output = "character"),
    parse = TRUE, ...
  )

}


```






# Step 1. Use the compute in a rendered plot

Sketch mode - where you don't actually define functions, we just have proposals commented out.


```{r prop_poem, echo = T, message = F, warning=F}
donor_data |>
  ggplot() +
  aes(x = decision) +
  # 1 geom_stack
  qlayer(geom = qproto_update(GeomTile, aes(color = from_theme(paper))), 
         stat = qstat(compute_group_bricks)) +
  # 2 geom_stack_label() 
  qlayer(geom = qproto_update(GeomText, aes(vjust = 0)), 
         stat = qstat(compute_group_count)) +
  # 3 geom_xrange, show scale, range at y is zero
  qlayer(geom = GeomSegment, 
         stat = qstat_panel(compute_balance)) +
  # 4. geom_prop, show prop, i.e. balancing point
  qlayer(geom = qproto_update(GeomText, aes(size = 6, vjust = 1)),
         stat = qstat_panel(compute_xmean_at_y0)) + 
  scale_x_discrete(palette = scales::pal_manual(0:1)) +
  # 5. geom_prop_label, labeling prop, balancing point
  qlayer(geom = qproto_update(GeomLabel, aes(fill = from_theme(colour %||% paper), label.size = NA, vjust = 0)),
         stat = qstat_panel(compute_xmean_at_y0_label))  +   
  # 6. stamp_prop, assertion, point
  qlayer(geom = qproto_update(GeomText, aes(size = 6, vjust = 1, color = from_theme(colour %||% accent))),
         stat = qstat_panel(compute_panel_prop_asserted), data = data.frame(x = 1), inherit.aes = F) +
  # 7. stamp_prop_label, assertion, label
  qlayer(geom = qproto_update(GeomLabel, 
                              aes(fill = from_theme(colour %||% paper), 
                                  label.size = NA, vjust = 0, 
                                  color = from_theme(colour %||% accent))),
         stat = qstat_panel(compute_panel_prop_asserted_label), data = data.frame(x = 1), inherit.aes = F) +
  stamp_eq_norm_prop() +
  # 8. geom_norm on prop plot
  qlayer(geom = qproto_update(GeomArea, aes(alpha = .2)),
         stat = qstat_panel(compute_dnorm_prop)) + 
   # 9. geom_prop_norm w/ sd marks
   qlayer(geom = qproto_update(GeomSegment, aes(linetype = "dotted")),
          stat = qstat_panel(compute_dnorm_prop_sds)) +
  labs(title = "Is there statistical evidence that choice to between being an\nan organ donar or not differs from 50/50") 
```

# And then define the functions...

<details>

```{r prop_poem_functions, echo = T, message = F, warning=F}


#' @export
geom_stack <- function(...){
  qlayer(geom = qproto_update(ggplot2::GeomTile, ggplot2::aes(color = "white")), 
         stat = qstat(compute_group_bricks), 
         ...)
  } 

#' @export
geom_stack_label <- function(...){
  qlayer(geom = qproto_update(ggplot2::GeomText, ggplot2::aes(vjust = 0)), 
         stat = qstat(compute_group_count), 
         ...)
  } 

#' @export
geom_support <- function(...){
  qlayer(geom = ggplot2::GeomSegment, 
         stat = qstat_panel(compute_balance), 
         ...)
  }


scale_x_prop <- function(...){ 

  scale_x_discrete(palette = scales::pal_manual(0:1), ...) 
  
}

#' @export
geom_prop <- function(...){
  list(
  qlayer(geom = qproto_update(ggplot2::GeomText, ggplot2::aes(size = 6, vjust = 1)),
         stat = qstat_panel(compute_xmean_at_y0),
         ...),
  scale_x_prop()
  )
  }

#' @export
geom_prop_label <- function(...){ 
  qlayer(geom = qproto_update(ggplot2::GeomLabel, 
                              ggplot2::aes(fill = ggplot2::from_theme(colour %||% paper), label.size = NA, vjust = 0)),
         stat = qstat_panel(compute_xmean_at_y0_label), 
         ...) 
  }

#' @export
stamp_prop <- function(...){ 
  qlayer(geom = qproto_update(ggplot2::GeomText, 
                              ggplot2::aes(size = 6, 
                                           vjust = 1, 
                                           color = ggplot2::from_theme(colour %||% accent))),
         stat = qstat_panel(compute_panel_prop_asserted), 
         data = data.frame(x = 1), 
         inherit.aes = FALSE,
         ...
         )
  }
  
#' @export  
stamp_prop_label <- function(...){  
  qlayer(geom = qproto_update(ggplot2::GeomLabel, 
                              ggplot2::aes(fill = ggplot2::from_theme(colour %||% paper), 
                                  label.size = NA, vjust = 0, 
                                  color = ggplot2::from_theme(colour %||% accent))),
         stat = qstat_panel(compute_panel_prop_asserted_label), 
         data = data.frame(x = 1), 
         inherit.aes = FALSE,
         ...
         )
  }

#' @export
geom_normal_prop_null <- function(...){
  qlayer(geom = qproto_update(ggplot2::GeomArea, ggplot2::aes(alpha = .2)),
         stat = qstat_panel(compute_dnorm_prop), 
         ...)
  } 

#' @export
geom_normal_prop_null_sds <- function(...){
   qlayer(geom = qproto_update(ggplot2::GeomSegment, ggplot2::aes(linetype = "dotted")),
          stat = qstat_panel(compute_dnorm_prop_sds), 
          ...)
  }

 


# #' @export
# geom_diff <- function(...){
#    qlayer(geom = qproto_update(ggplot2::GeomSegment, ggplot2::aes(linetype = "dotted")),
#           stat = qstat_panel(compute_dnorm_prop_sds), 
#           ...)
#   
#      qlayer(geom = qproto_update(ggplot2::GeomSegment, ggplot2::aes(linetype = "dotted")),
#           stat = qstat_panel(compute_dnorm_prop_sds), 
#           ...)
#   
#   }




```

</details>


# Done!  See below for all layers, and the introduction of packag for potential break-points along the prop test train of thought.

```{r}
library(ggplot2)
library(ggprop.test)
library(S7)

theme_set(
  ggchalkboard:::theme_blackboard(
    base_size = 12))


layers_remove <- function(i = NULL) {

  structure(
    list(index_wipe = i), 
    class = "layers_remove"
    )

}


method(update_ggplot, list(new_S3_class("layers_remove"), class_ggplot)) <- 
  function(object, plot, ...) {
    
  if(is.null(object$index_wipe)){
   
    plot$layers <- list()
     
  }else{
    
    plot$layers[object$index_wipe] <- NULL

  }
  
  plot
    
  }


donor_data |>
  ggplot() + 
  aes(x = decision) +
  geom_stack() + 
  geom_stack_label() + 
  geom_support() + 
  geom_prop() + 
  geom_prop_label() + 
  stamp_prop() + 
  stamp_prop_label() + 
  geom_normal_prop_null() +
  geom_normal_prop_null_sds() + 
  stamp_eq_norm_prop()


```

</details>



```{r}
# ?prop.test
prop.test(x = sum(donor_data$Choice == "donor"), # n true
          n = length(donor_data$Choice), # num observations
          p = .5) # null
```




---

```{r, eval = F, echo = F}
stamp_probs <- function(prob = .5, ...){
  
  data = data.frame(x = 0:1 |> as.factor(), y = c(1-prob, prob), xend = 0:1 |> as.factor(), yend = 0)
  
  list(
  geom_segment(data = data, aes(x = x, y =y, xend = xend, yend = yend), inherit.aes = F, ...),
  geom_point(data = data, aes(x = x, y =y), inherit.aes = F,...)
  )
  
}

stamp_support <- function(xmin = 0, xmax = 1, ...){
  
    data = data.frame(x = xmin |> as.factor(), y = 0, 
                      xend = xmax |> as.factor(), yend = 0)

  
    geom_support(data = data, aes(x = x, y = y, xend = xend, yend = yend), inherit.aes = F, ...)

  
}


stamp_probs_sample <- function(prob = .5, n, ...){

  x = sample(x = 0:1 |> as.factor(), size = n, replace = T)
  
  data <- data.frame(x = x)
  
  geom_stack(data = data, aes(x = x), ..., inherit.aes = F)

  
}



```


```{r, eval = T}

plausible_based_on_null <- function(data, var, prob = .5){
  
  observed <- data |> 
    pull({{var}})
  
  generated <- levels(observed) |>  # take two 
    sample(size = nrow(data), replace = T) |> 
    # restore category ordering
    factor(levels = levels(observed))
  
  data |> 
    mutate(plausible = generated)
  
}
```

```{r}
plausible_based_on_null(donor_data, var = decision) 

donor_data |> 
  plausible_based_on_null(decision) |>
  ggplot() + 
  aes(x = plausible) + 
  labs(x = "plausible balance if null (50-50) is true") +
  geom_stack() + 
  geom_prop() + 
  geom_prop_label() + 
  geom_support() + 
  stamp_prop(value = .67)

last_plot() + plausible_based_on_null(donor_data, var = decision) 
last_plot() + plausible_based_on_null(donor_data, var = decision) 
last_plot() + plausible_based_on_null(donor_data, var = decision) 
last_plot() + plausible_based_on_null(donor_data, var = decision) 
last_plot() + plausible_based_on_null(donor_data, var = decision) 
last_plot() + plausible_based_on_null(donor_data, var = decision) 
last_plot() + plausible_based_on_null(donor_data, var = decision) 
```


---

# Minimal Packaging


```{r}
knitrExtra::chunk_to_dir("statexpress")
knitrExtra::chunk_to_dir("compute_for_prop_story")
knitrExtra::chunk_to_dir("prop_poem_functions")
knitrExtra::chunk_to_dir("stamp_eq_norm_prop")
```

```{r, eval = F}
usethis::use_package("dplyr")
usethis::use_package("ggplot2")

devtools::document()
devtools::check(".")
devtools::install(pkg = ".", upgrade = "never")
```


<!-- # epilogue... -->


<!-- Another approach that is less concerned with being a bridge to ggplot2 layer extension...?   -->

<!-- ```{r, eval = F, echo = F} -->
<!-- geom_support <- function(...){geom_segment(data = compute_balance, ...)} -->
<!-- geom_prop <- function(...){geom_point(data = compute_xmean_at_y0, label = "^", ...)} -->

<!-- donor |> -->
<!--   ggplot() +  -->
<!--   aes(x = decision) + -->
<!--   geom_support() + -->
<!--   geom_prop() -->
<!-- ``` -->
